---
title: "Compare TADs with del. enhancer to normal TADs"
output: html_document
---

> Compile with `Rscript -e 'rmarkdown::render("ase_genes_in_TADs_with_DEL_enhancers.Rmd", clean=F, output_format="all");'`


```{r, echo=F,message=FALSE, warning=FALSE}
source("common.R")
```


# Input data

The following variables are read by `common.R`:

  * `ASE.adult`, `ASE.embryo`
  * `EXONS`, `GENE_SPAN`, `GENE_TSS`


### Read TAD calls

```{r}
TAD_SEXTON = fread("../../TADs/data/TADs_Sexton_2012_embryo_16-18h_dm6.bed", sep="\t", select = 1:3)
TAD_SEXTON = TAD_SEXTON[grepl('^chr[23][LR]$',V1),]
TAD_OURS   = fread("../../TADs/data/TADs_HiC_DB_6-8h_All_IS100k_dm6.bed", sep="\t", select = 1:3)
TAD_OURS   = TAD_OURS[grepl('^chr[23][LR]$',V1),]
```


### Read deletions and enhancers

```{r}
DEL_BAL = fread("../../tracks/deletions/min20.merged.bal-spec.bed")
DEL_VRG = fread("../../tracks/deletions/min20.merged.vrg-spec.bed")
ENH_CAD = fread("../../tracks/enhancer/CAD4_plus_vienna_dm6.core80percent.onlyChr2and3.bed")
ENH_DHS = fread("../../tracks/enhancer/macs_summits_p50_slop50_merge25_clean.htseq.dm6.onlyChr2and3.bed")
```




# Gene counts in TADs

I take 4 different (potentially overlapping) subsets of TADs, depending on

  * whether they contain a **DHS** summit or a **CAD4** region
  * which is hit by a **bal-spec** or a **vrg-spec** deletion

~~As a comparison I use the total set of sexton TAD calls.~~

No, I will randomly draw TADs. that gets us a little bit closer to the selected TADs in terms of size and # tested genes. Again I need to remove centromers for that:

```{r}
manual = GRanges(c("chr2L",     "chr2R",        "chr3L",        "chr3R"),
        IRanges(c(1000,         5e6,            1000,           4.2e6),
                c(23513712-5e5, 25286936 - 1e3, 28110227-5.3e6, 32079331 - 1e3)))
```

```{r}
compare_TADs <- function(dt, bg) {
    dt.genes <- as.data.table(get_genes_in_region(GENE_TSS, GRanges(dt[[1]], IRanges(dt[[2]],dt[[3]]))))
    bg.genes <- as.data.table(get_genes_in_region(GENE_TSS, GRanges(bg[[1]], IRanges(bg[[2]],bg[[3]]))))
    d <- rbind(cbind(dt.genes, group="test"), cbind(bg.genes, group="bg"))
    d$size = d$end - d$start
    d.m <- melt.data.table(d, id.vars = "group", measure.vars = c("n_ase.em","n_ase.ad","n_tested.em", "n_tested.ad","size"))
    plt <- ggplot(d.m) + aes(x=group,y=value) + geom_boxplot() + geom_jitter(alpha=0.3) + facet_wrap(~variable, scales="free")
    return(plt)
}
```

The following plots show number of measured and ASE genes for affected TADs vs. other TADs:

  * `n_ase.*` shows the number of ASE genes
  * `n_tested.*` shows the number of tested genes
  
```{r genes_affected_in_TADs_DHS_bal, fig.width=6, fig.height=6, ev=c('png','pdf')}
x1 <- subsetByOverlapsDT(TAD_SEXTON, subsetByOverlapsDT(ENH_DHS, DEL_BAL, 0.1))
#x0 <- TAD_SEXTON[!(paste0(V1,V2,V3) %in% paste0(x1$seqnames,x1$start,x1$end)),]
x0 <- subsetByOverlapsDT(TAD_SEXTON, as.data.table(get_random_breakpoints(200, manual)))
compare_TADs(x1, x0) + 
  ggtitle(paste0("TADs with del-spec DHS affected (n=",nrow(x1),") vs. other TADs (n=",nrow(x0),")"))
```

```{r genes_affected_in_TADs_DHS_vrg, fig.width=6, fig.height=6, ev=c('png','pdf')}
x3 <- subsetByOverlapsDT(TAD_SEXTON, subsetByOverlapsDT(ENH_DHS, DEL_VRG, 0.1))
#x2 <- TAD_SEXTON[!(paste0(V1,V2,V3) %in% paste0(x3$seqnames,x3$start,x3$end)),]
compare_TADs(x3, x0) + 
  ggtitle(paste0("TADs with vrg-spec DHS affected (n=",nrow(x3),") vs. other TADs (n=",nrow(x0),")"))
```

```{r genes_affected_in_TADs_CAD_bal, fig.width=6, fig.height=6, ev=c('png','pdf')}
x5 <- subsetByOverlapsDT(TAD_SEXTON, subsetByOverlapsDT(ENH_CAD, DEL_BAL, 0.1))
#x4 <- TAD_SEXTON[!(paste0(V1,V2,V3) %in% paste0(x5$seqnames,x5$start,x5$end)),]
compare_TADs(x5, x0) + 
  ggtitle(paste0("TADs with del-spec CAD4 affected (n=",nrow(x5),") vs. other TADs (n=",nrow(x0),")"))
```

```{r genes_affected_in_TADs_CAD_vrg, fig.width=6, fig.height=6, ev=c('png','pdf')}
x7 <- subsetByOverlapsDT(TAD_SEXTON, subsetByOverlapsDT(ENH_CAD, DEL_VRG, 0.1))
#x6 <- TAD_SEXTON[!(paste0(V1,V2,V3) %in% paste0(x7$seqnames,x7$start,x7$end)),]
compare_TADs(x7, x0) + 
  ggtitle(paste0("TADs with vrg-spec CAD4 affected (n=",nrow(x7),") vs. other TADs (n=",nrow(x0),")"))
```


# Candidates

Here is my **strict** selection:

```{r}
candidates <- unique(subsetByOverlapsDT(TAD_SEXTON, 
                rbind(subsetByOverlapsDT(ENH_DHS, DEL_BAL, 0.1),
                      subsetByOverlapsDT(ENH_DHS, DEL_VRG, 0.1))) )
```

# General info on TADs

```{r}
TAD_SEXTON_GENES <- get_genes_in_region(GENE_TSS, with(TAD_SEXTON, GRanges(V1, IRanges(V2,V3))))
```

```{r tad_size_vs_tested_genes, fig.width=8, fig.height=6, ev=c('png','pdf')}
E <- melt.data.table(as.data.table(TAD_SEXTON_GENES), 
                     c("seqnames", "start", "end"), 
                     c("n_tested.ad", "n_tested.em"))
E$variable <- factor(E$variable, levels=c("n_tested.ad", "n_tested.em"), labels=c("adult","embryo"))
ggplot(E) + aes(end - start, value) + 
  geom_point(alpha=0.5) + 
  scale_x_log10(labels=comma) + 
  geom_smooth(method = "loess", col="red", level=0.99) + 
  facet_grid(variable~.) +
  xlab("TAD size") + 
  ylab("Number of tested genes") +
  annotation_logticks(sides = "b") +
  ggtitle(paste0("Sexton TADs (n=", length(TAD_SEXTON_GENES), ")"))
```

## Plot all TADs and see where the candidates fall:

```{r tad_number_genes_vs_aseratio, fig.width=8, fig.height=6, ev=c('png','pdf')}
E <- melt.data.table(as.data.table(TAD_SEXTON_GENES), 
                     c("seqnames", "start", "end", "n_tested.ad", "n_tested.em"), 
                     c("r_ase.ad", "r_ase.em"),
                     variable.name = "sample",
                     value.name = "ratio")
E$sample <- factor(E$sample, levels=c("r_ase.ad", "r_ase.em"), labels=c("adult","embryo"))
E <- melt.data.table(E, c("seqnames", "start", "end", "sample", "ratio"),
                variable.name = "sample2",
                value.name = "n_tested")
E$sample2 <- factor(E$sample2, levels=c("n_tested.ad", "n_tested.em"), labels=c("adult","embryo"))
E <- E[sample == sample2, ]

# Add information about candidates:
E$candidate = paste(E$seqnames, E$start, E$end) %in% paste(candidates$V1, candidates$V2, candidates$V3)

ggplot(E) + aes(n_tested, ratio) + 
  stat_sum(aes(size=..n..), col='darkgrey') +
  geom_point(col = 'darkgrey') + 
  facet_grid(sample~.) +
  xlab("Number of tested genes") + 
  ylab("Ratio of ASE genes") + 
  geom_point(data = E[E$candidate,], col="dodgerblue") +
  ggtitle("Ratio of ASE genes in Sexton TADs")
```

Plot a box plot of ASE ratio only for TADs with at least 2 expressed genes.

```{r violin_aseratio, fig.width=6, fig.height=3, ev=c('png','pdf')}
MIN_TESTED=5
table(E[n_tested>=MIN_TESTED,]$candidate, E[n_tested>=MIN_TESTED,]$sample)
ggplot(E[n_tested>=MIN_TESTED & ratio>0,]) + aes(candidate, ratio) + geom_boxplot() + facet_grid(.~sample) + xlab("has 10% enhancer deletion") + ylab("Ratio of ASE genes") + geom_jitter(alpha=0.3) + ggtitle(paste("TADs with at leasn", MIN_TESTED, "expr. genes"))
```

```{r violin_number_tested, fig.width=6, fig.height=3, ev=c('png','pdf')}
MIN_TESTED=10
table(E[n_tested>=MIN_TESTED,]$candidate, E[n_tested>=MIN_TESTED,]$sample)
ggplot(E[n_tested>=MIN_TESTED & ratio>0,]) + aes(candidate, ratio) + geom_boxplot() + facet_grid(.~sample) + xlab("has 10% enhancer deletion") + ylab("Ratio of ASE genes") + geom_jitter(alpha=0.3) + ggtitle(paste("TADs with at leasn", MIN_TESTED, "expr. genes"))
```

Another way to plot it:

```{r tad_number_genes_vs_number_ase, fig.width=8, fig.height=6, ev=c('png','pdf')}
E <- melt.data.table(as.data.table(TAD_SEXTON_GENES), 
                     c("seqnames", "start", "end", "n_tested.ad", "n_tested.em"), 
                     c("n_ase.ad", "n_ase.em"),
                     variable.name = "sample",
                     value.name = "n_ase")
E$sample <- factor(E$sample, levels=c("n_ase.ad", "n_ase.em"), labels=c("adult","embryo"))
E <- melt.data.table(E, c("seqnames", "start", "end", "sample", "n_ase"),
                variable.name = "sample2",
                value.name = "n_tested")
E$sample2 <- factor(E$sample2, levels=c("n_tested.ad", "n_tested.em"), labels=c("adult","embryo"))
E <- E[sample == sample2, ]

# Add information about candidates:
E$candidate = paste(E$seqnames, E$start, E$end) %in% paste(candidates$V1, candidates$V2, candidates$V3)

ggplot(E) + aes(n_tested, n_ase) + 
  stat_sum(aes(size=..n..), col='darkgrey') +
  geom_point(col = 'darkgrey') + 
  facet_grid(sample~.) +
  xlab("Number of tested genes") + 
  ylab("Number of ASE genes") + 
  geom_point(data = E[E$candidate,], col="dodgerblue") +
  ggtitle("Number of ASE genes in Sexton TADs")
```

Here are the potentially intersting TADs (top 20):

```{r}
d <- E[candidate & n_ase>0,]
d$sample2 = d$candidate = NULL
setorder(d, -n_ase, n_tested)
head(d, n=20)
```
