---
title: "ASE analysis comparison"
author: "Sascha Meiers"
date: "May 25, 2016"
output: html_document
---

> Compile with `Rscript -e 'library(knitr); rmarkdown::render("post_analysis.Rmd", clean=F, output_format="all");'`

# Comparing new and old adult F1 data

```{r}
source("common.R")
suppressPackageStartupMessages(library(data.table))
DE_newF1adult = fread("DESeq.F1_headsFemale.standardFormat.txt")
DE_oldF1adult = fread("DESeq.F1_heads.standardFormat.txt")
DE_allF1adult = merge(DE_newF1adult[,.(gene_id, log2FoldChange, pvalue, padj)], 
                      DE_oldF1adult[,.(gene_id, log2FoldChange, pvalue, padj)], 
                      by="gene_id", suffixes=c(".new",".old"))
```

```{r adult_significant_venn, fig.width=4, fig.height=3}
t = DE_allF1adult[, .(c = length(gene_id)), by = .(sign.new = padj.new < 0.05, sign.old = padj.old < 0.05)]
library(VennDiagram)
grid.newpage()
draw.pairwise.venn(area1 = sum(t$c[t$sign.old==T]), 
                   area2 = sum(t$c[t$sign.new==T]), 
                   cross.area = sum(t$c[t$sign.old==T & t$sign.new==T]), 
                   category = c("Old", "New"),
                   lty = rep("blank", 2), fill = c("skyblue", "orange"),
                   alpha = rep(0.4, 2), euler.d = TRUE, rotation.degree = 45)
```

```{r adultF1_significant_lfc_corr, fig.width=7, fig.height=4}
ggplot(DE_allF1adult[padj.new < 0.05 | padj.old < 0.05,]) + 
  aes(log2FoldChange.new, log2FoldChange.old, col=interaction(padj.old < 0.05, padj.new < 0.05)) + 
  geom_point(alpha=0.7) + 
  scale_color_discrete(name = "Sign. (old, new)") +
  coord_cartesian(xlim=c(-3,3), ylim=c(-3,3), expand=F) +
  scale_x_continuous(breaks=-2:2) +
  scale_y_continuous(breaks=-2:2) +
  ggtitle("Significant ASE genes in old and new RNA")
```



# Comparing (old) adult and embryonic RNA

```{r adult_embryo_significant_venn, fig.width=4, fig.height=3}
DE_N1embryo     = fread("DESeq.N1_6-8h.standardFormat.txt")
DE_oldF1adult = fread("DESeq.F1_heads.standardFormat.txt")
DE_em_ad = merge(DE_N1embryo[,.(gene_id, log2FoldChange, pvalue, padj)], 
                 DE_oldF1adult[,.(gene_id, log2FoldChange, pvalue, padj)], 
                 by="gene_id", suffixes=c(".embryo",".adult"))
t = DE_em_ad[, .(c = length(gene_id)), by = .(sign.embryo = padj.embryo < 0.05, sign.adult = padj.adult < 0.05)]
grid.newpage()
draw.pairwise.venn(area1 = sum(t$c[t$sign.embryo==T]), 
                   area2 = sum(t$c[t$sign.adult==T]), 
                   cross.area = sum(t$c[t$sign.embryo==T & t$sign.adult==T]), 
                   category = c("Embryo", "Adult (old)"),
                   lty = rep("blank", 2), fill = c("skyblue", "orange"),
                   alpha = rep(0.4, 2), euler.d = TRUE, rotation.degree = 45)
```

```{r adultembryo_significant_lfc_corr, fig.width=7, fig.height=4}
ggplot(DE_em_ad[padj.embryo < 0.05 | padj.adult < 0.05,]) + 
  aes(log2FoldChange.embryo, log2FoldChange.adult, col=interaction(padj.embryo < 0.05, padj.adult < 0.05)) + 
  geom_point(alpha=0.7) + 
  scale_color_discrete(name = "Sign. (embryo, adult)") +
  coord_cartesian(xlim=c(-3,3), ylim=c(-3,3), expand=F) +
  scale_x_continuous(breaks=-2:2) +
  scale_y_continuous(breaks=-2:2) +
  ggtitle("Significant ASE genes in embryos or (old) adult data")
```

----

# Old 

## Genes with decent coverage (in embryos)

```{bash}
#comm -1 -2 <(cut -f5 ../../gatk_asereadcounter/stats/N1_pool_6-8h.SNVs_balancers.SNVs.txt | sort -u) \
#           <(cut -f5 ../../gatk_asereadcounter/stats/N1sex_pool_6-8h.SNVs_balancers.SNVs.txt | sort -u) \
#           > N1_6-8.gene_ids.txt
```



# Comparison of adult and embryonic data

```{r}
# adult = read.table("DESeq.adult.chr2_3.txt", header=T) %>%
#     mutate(fpkm  = (F1_norm.1.fpkm + F1_norm.2.fpkm + N1_norm.1.fpkm + N1_norm.2.fpkm)/4,
#            sample = "adult", significant = cis.padj <= 0.05) %>%
#     select(-starts_with("F1"), -starts_with("N1"), -starts_with("cis"), 
#            padj = cis.padj, lfc = cis.lfc)
# 
# embryo = read.table("DESeq.N1_6-8h.chr2_3.txt", header=T) %>%
#     mutate(fpkm  = (N1_norm.1.fpkm + N1_norm.2.fpkm + N1_rsex.1.fpkm + N1_rsex.2.fpkm)/4,
#            sample = "embryo", significant = cis.padj <= 0.05) %>%
#     select(-starts_with("N1"), -starts_with("int"), -starts_with("unf"), -starts_with("cis"),
#            -length, padj = cis.padj, lfc = cis.lfc)
#     
# long = rbind(adult, embryo)
# wide = merge(adult, select(embryo, -from, -to, -strand), by="gene_id", 
#              suffixes = c(".adult", ".embryo"))
```

Correlation of total gene expression is not very high between adult fly heads
and 6-8h embryos:

```{r, fig.width=6, fig.height=6, dev=c('png','pdf')}
# ggplot(wide) + aes(fpkm.adult, fpkm.embryo) + 
#   geom_point(alpha=0.2) + 
#   scale_x_log10() + scale_y_log10() + 
#   ggtitle("Gene expression") +
#   theme_minimal()
```

```{r, fig.width=6, fig.height=6, dev=c('png','pdf')}
# subset <- wide %>% filter(fpkm.adult >= 2, fpkm.embryo >= 2)
# ggplot(subset) + aes(lfc.adult, lfc.embryo) + geom_point(alpha=0.2) + 
#   #coord_cartesian(xlim=c(-2,2), ylim=c(-2,2)) +
#   xlab("embryonic balancer/vrg lfc") + 
#   ylab("adult balancer/vrg lfc") + 
#   ggtitle(paste(dim(subset)[1], "ASE for genes expressed in both")) + 
#   theme_minimal()
```

```{r, fig.width=6, fig.height=6, dev=c('png','pdf')}
# ggplot(subset) + aes(lfc.embryo, lfc.adult, 
#                   col = interaction(padj.embryo <= 0.05, padj.adult <= 0.05)) + 
#   geom_point(alpha=0.4) + theme_minimal() + 
#   scale_color_discrete(name = "Significance: embryo.adult") + 
#   theme(legend.position = c(0, 1), legend.justification = c(0, 1)) + 
#   ggtitle("Significant ASE in embryos and adults")
# table(with(subset,interaction(padj.embryo <= 0.05, padj.adult <= 0.05)))
```

Write down the list of common genes

```{r}
# subset %>% mutate(embryo.adult = interaction(padj.embryo <= 0.05, padj.adult <= 0.05)) %>%
#   select(gene_id, embryo.adult, lfc.embryo, lfc.adult) %>%
#   arrange(embryo.adult, gene_id) %>% 
#   write.table(., file = "DESeq.comparison.txt", quote=F, row.names=F, sep="\t")
```

### Sanity checks

```{r}
# ggplot(long) + aes(significant, fpkm) + geom_violin() + 
#   geom_jitter(size=0.2, alpha=0.2) + facet_grid(.~sample) +
#   scale_y_log10() + theme_minimal() + ggtitle("Significance vs. expression")
```

```{r}
# ggplot(long) + aes(significant, to-from) + geom_violin() + 
#   geom_jitter(size=0.2, alpha=0.2) + facet_grid(.~sample) +
#   scale_y_log10() + theme_minimal() + ggtitle("Significance vs. size")
```

```{r}
# ggplot(long) + aes(significant, from) + geom_violin(adjust = .5) + 
#   geom_jitter(size=0.2, alpha=0.2) + facet_grid(.~sample) +
#   theme_minimal() + ggtitle("Significance vs. position")
```

## Write bed files

```{r}
# embryo %>% select(chrom, from, to, gene_id, significant) %>% 
#   arrange(chrom, from) %>%
#   write.table(. ,"bed/embryo.bed", quote = F, row.names = F, col.names = F, sep = "\t")
# adult %>% select(chrom, from, to, gene_id, significant) %>% 
#   arrange(chrom, from) %>%
#   write.table(. ,"bed/adult.bed", quote = F, row.names = F, col.names = F, sep = "\t")
```


## Test area - this depends on some files I created temporarily (bedtools)

This has to be re-written correctly and systematically

```{r}
# d = read.table("bed/test.adult.LargeBPdist.txt")
# ggplot(d) + aes(V5, V10) + geom_boxplot() + geom_jitter(size = 0.2, alpha = 0.2) + ggtitle("Distance to large INV breakponts (adult)")
# t.test(d$V10[d$V5], d$V10[!d$V5]) 
```

```{r}
# d = read.table("bed/test.adult.DELdist.vrg-spec.txt") %>%
#     select(gene_id = V4, chrom2 = V6, start2 = V7, end2 = V8, dist = V9) %>%
#     merge(., adult, by = "gene_id") %>%
#     mutate(lfc.fac = factor(lfc>0, levels=c(F,T), labels=c("balancer down","balancer up")))
# ggplot(d) + aes(significant, dist) + geom_boxplot() + 
#     geom_jitter(size = 0.2, alpha = 0.2) + scale_y_log10() + facet_wrap(~lfc.fac) + 
#     ggtitle("Distance to vrg-spec. DELs  (adult)")
```

```{r}
# d = read.table("bed/test.adult.DELdist.bal-spec.txt") %>%
#     select(gene_id = V4, chrom2 = V6, start2 = V7, end2 = V8, dist = V9) %>%
#     merge(., adult, by = "gene_id") %>%
#     mutate(lfc.fac = factor(lfc>0, levels=c(F,T), labels=c("balancer down","balancer up")))
# ggplot(d) + aes(significant, dist) + geom_boxplot() + 
#     geom_jitter(size = 0.2, alpha = 0.2) + scale_y_log10() + facet_wrap(~lfc.fac) + 
#     ggtitle("Distance to bal-spec. DELs  (adult)")
```
